Diverse Genotype-by-Weather Interactions in Switchgrass

### 2024-06-10 R/Analysis_v1.2_mash_using_greedy_covar.qmd
revision2: edit M&M based on Tom's March comments and new SI

ASReml h2 greenup analysis
```{r}
#| eval: false
#| 
library(tidyverse)
library(asreml)
library(switchgrassGWAS)
library(cowplot)

@@

get BLUPs from ASReml
```{r}
site_v <- list(eight_sites = c("TX1", "TX2", "TX3", "OK", "MO", "NE", "MI", "SD"))
subpop_v <- list(Gulf = "Gulf", Midwest = "Midwest", Gulf_and_Midwest = c("Gulf", "Midwest"))
phe_fl <- c("FL50", "dyln_fl50", "dyln_change_sec", "cgdd_12c_gr2fl")
phe_gr <- c("GR50", "tmin_10d", "cgdd_12c_10d")
phe_dfs <- list(phe_gr_df, phe_fl_df)
phe_lists <- list(phe_gr, phe_fl)
h2_table <- tibble()

@@

set up phenotype df for gwas
```{r}
#| eval: false 
#| 
site_v <- list(eight_sites = c("TX1", "TX2", "TX3", "OK", "MO", "NE", "MI", "SD"))
subpop_v <- list(Gulf = "Gulf", Midwest = "Midwest", Gulf_and_Midwest = c("Gulf", "Midwest"))
phe_fl <- c("FL50", "dyln_fl50", "dyln_change_sec", "cgdd_12c_gr2fl")
phe_gr <- c("GR50", "tmin_10d", "cgdd_12c_10d")
phe_dfs <- list(phe_gr_df, phe_fl_df)
phe_lists <- list(phe_gr, phe_fl)

@@

NB: only the single_site BLUPs & GWAS that I generated were used in the subsequent analysis.
```{r}
#| eval: false
#| 
library(bigsnpr)
library(tidyverse)
library(switchgrassGWAS)
library(AnnotationDbi)
library(here)
txdb <- loadDb(file = here("data-raw", "ignore",
                           "Pvirgatum_516_v5.1.gene.txdb.sqlite"))
workingdir <- file.path("~", "Github", "pvdiv-phenology-gxe")
inputdir <- file.path(workingdir, "analysis", "gwas", "BLUPs")
outputdir <- file.path(workingdir, "analysis", "gwas", "BLUPs")
subpop_v <- list(Gulf_and_Midwest = c("Gulf", "Midwest"), Midwest = "Midwest", Gulf = "Gulf")
snp_v <- c("12\\.1M", "8\\.8M", "10\\.3M")

@@

# ---
# Step 5: Redo writeup & figures

>>>>

ASReml h2 greenup analysis

```{r}
#| eval: false
#| 
library(tidyverse)
library(asreml)
library(switchgrassGWAS)
library(cowplot)

@@

get BLUPs from ASReml

```{r}
site_v <- list(eight_sites = c("TX1", "TX2", "TX3", "OK", "MO", "NE", "MI", "SD"))
subpop_v <- list(Gulf = "Gulf", Midwest = "Midwest", Gulf_and_Midwest = c("Gulf", "Midwest"))
phe_fl <- c("FL50", "dyln_fl50", "dyln_change_sec", "cgdd_12c_gr2fl")
phe_gr <- c("GR50", "tmin_10d", "cgdd_12c_10d")
phe_dfs <- list(phe_gr_df, phe_fl_df)
phe_lists <- list(phe_gr, phe_fl)
h2_table <- tibble()

@@

set up phenotype df for gwas

```{r}
#| eval: false 
#| 
site_v <- list(eight_sites = c("TX1", "TX2", "TX3", "OK", "MO", "NE", "MI", "SD"))
subpop_v <- list(Gulf = "Gulf", Midwest = "Midwest", Gulf_and_Midwest = c("Gulf", "Midwest"))
phe_fl <- c("FL50", "dyln_fl50", "dyln_change_sec", "cgdd_12c_gr2fl")
phe_gr <- c("GR50", "tmin_10d", "cgdd_12c_10d")
phe_dfs <- list(phe_gr_df, phe_fl_df)
phe_lists <- list(phe_gr, phe_fl)

@@

NB: only the single_site BLUPs & GWAS that I generated were used in the subsequent analysis.

```{r}
#| eval: false
#| 
library(bigsnpr)
library(tidyverse)
library(switchgrassGWAS)
library(AnnotationDbi)
library(here)
txdb <- loadDb(file = here("data-raw", "ignore",
                           "Pvirgatum_516_v5.1.gene.txdb.sqlite"))
workingdir <- file.path("~", "Github", "pvdiv-phenology-gxe")
inputdir <- file.path(workingdir, "analysis", "gwas", "BLUPs")
outputdir <- file.path(workingdir, "analysis", "gwas", "BLUPs")
subpop_v <- list(Gulf_and_Midwest = c("Gulf", "Midwest"), Midwest = "Midwest", Gulf = "Gulf")
snp_v <- c("12\\.1M", "8\\.8M", "10\\.3M")

@@

# ---

# Step 5: Redo writeup & figures

### 2024-06-10 manuscript/revision2/SI_Appendix/SI_Appendix.qmd
revision2: edit M&M based on Tom's March comments and new SI

## Section S1. Covariance matrices generated to provide multiple ways SNP effects could covary across common gardens {.unnumbered}

A trait measured in two environments can be considered two traits that are genetically correlated [@falconer1952]. By extension, a phenological trait measured at eight common gardens can be modeled as eight genetically correlated traits represented by a covariance matrix, where diagonal entries represent narrow-sense heritability at a common garden, or the additive genetic variance present in the population at that garden, and non-diagonal entries represent the genetic correlation between the trait in a pair of environments.

We were interested in the types of genetic correlation that could be expressed by different alleles across the genome and reasoned that different loci could have different patterns of genetic correlation. For example, the effects of a locus, $A$, could covary with a weather-based cue if $A$ had a common allele, $a$, that was responsive to this cue. In contrast, the effects of a different locus $B$ might have an effect at only one garden, for example if a pathogen was only present at one garden, and $B$ has a common allele $b$ that is resistant to that pathogen.

@@

To create the GxWeather covariance matrices we transformed each genotype's green-up and flowering date at each garden to create date-related values that were summary functions for a weather variable across a specified date range. For example, the cumulative sum of the rainfall that occurred seven days prior to the flowering date was used to create one possible GxWeather cue variable for flowering date. The code for these calculations can be found at git\@github.com:Alice-MacQueen/pvdiv-phenology-gxe/R/Analysis_v1.2_mash_using_greedy_covar.qmd, starting at line 390.

@@

We then generated GxWeather covariance matrices derived from correlations in these GxWeather cue variables for the three subpopulations. These covariance matrices represent the correlations between genotypes for these weather-related variables across our eight common gardens. To create these matrices, we determined the correlations between these weather-related cue variables for identical genotypes grown at different common gardens and used these correlations to fill the off-diagonals of a eight-by-eight matrix. To fill the diagonal entries of these matrices, we used the narrow-sense heritability for each weather-related cue variable at each garden, calculated using the same methods as Section S3. Matrices need to be positive semi-definite to be used in *mash*, so when the resultant matrices were not positive semi-definite, the `Matrix::nearPD` function was used to compute the nearest positive definite matrix to the correlation matrix [@Matrix]. Finally, all covariance matrices were rescaled so that the maximum diagonal element of the rescaled matrix is 1, as in [@Urbut2019]. Code used to generate these matrices can be found at git\@github.com:Alice-MacQueen/pvdiv-phenology-gxe.git/R/Analysis_v1.2_mash_using_greedy_covar.qmd, starting at line 716.

@@

To obtain initial effect estimates for each single nucleotide polymorphism (SNP) for each combination of phenological trait, genetic subpopulation, and common garden, we conducted univarate genome-wide association (GWAS) on site-specific best linear unbiased predictors (BLUPs) using the switchgrassGWAS package and the methods in [@Lovell2021]. The full SNP diversity panel generated in [@Lovell2021] was subset within each subpopulation for all subsequenct analyses to retain only SNPs with $\leq$ 20% missing data and minor allele frequencies \> 0.05. This subsetting resulted in 8.8 million SNPs retained for the Midwest subpopulation, 10.3 million SNPs retained for the Gulf subpopulation, and 12.3 million SNPs retained for Both subpopulations.

@@

where the vector $\mathbf{y}$ represents the flowering date or greenup date values for that garden, $Z$ the design matrix for random effects, $u$ the whole genome additive genetic effect, $e$ the residual, $G$ the genomic relationship matrix based on all SNPs retained for subpopulation-specific analyses, and $I$ the rank-$\mathbf{y}$ identity matrix [@clark2013]. ASReml can specify considerably more complex variance structures; however, we were interested in specifying covariance structures using *mash*, which allows the specification of multiple covariance structures; hence, we used a relatively simple model to calculate BLUPs for use in univariate GWAS. The code for these calculations can be found at git\@github.com:Alice-MacQueen/pvdiv-phenology-gxe.git/R/Analysis_v1.2_mash_using_greedy_covar.qmd, starting at line 74.

Univariate GWAS were run independently for each combination of trait, genetic subpopulation, and common garden. The code used to run GWAS can be found at git\@github.com:Alice-MacQueen/pvdiv-phenology-gxe.git/R/Analysis_v1.2_mash_using_greedy_covar.qmd, starting at line 240. We used [`pvdiv_standard_run()`](https://alice-macqueen.github.io/switchgrassGWAS/reference/pvdiv_standard_gwas.html), an opinionated function of the [switchgrassGWAS](https://github.com/Alice-MacQueen/switchgrassGWAS) R package, to conduct GWAS. To control for population structure and run GWAS, this function uses a three-part process. First, a fast truncated singular value decomposition (SVD) is used, which has initial pruning of regions in long-range linkage disequilibrium (LD) and iterative pruining and removal of regions in long-range LD [@privé2017a]. By default, 15 vectors of singular values, $k$, are computed. Pruning and removal of long-range LD regions are recommended in human genetics to best control for population structure [@abdellaoui2013; @privé2017; @price2008]. Second, linear regression models using 0-15 of these vectors of singular values were used to run GWAS and to calculate the genomic control coefficient, $\lambda_{GC}$, for each $k$ [@Devlin1999]. Third, we chose the lowest value of $k$ that made $\lambda_{GC}$ closest to 1 to correct for population structure, and ran GWAS using a linear regression model with this $k$ for population structure correction.

@@

To make the subsequent analyses using *mash* computationally feasible, we followed the author's [recommended analysis outline](https://stephenslab.github.io/mashr/articles/eQTL_outline.html) for large datasets using *mash* and considered subsets of the millions of effects at different stages of the mash analysis. We extracted two subsets of tests from our complete GWAS effect estimate datasets: 1) effects from a subset of "strong" tests corresponding to stronger effects on our traits; 2) results from a random subset of all tests to correspond to an unbiased representation of all effects. We also clumped effects before selection so that signal strength was standardized across different LD blocks in the genome, and not limited to closely linked associations.

@@

## Section S4. Greedy *mash* algorithm used to select covariance matrices that significantly improved the model log-likelihood {.unnumbered}

Because we were particularly interested in determining whether genomic loci varied in their patterns of genetic covariance across gardens, specifying the optimal set of covariance matrices in *mash* was an important model-building problem. If very similar covariance matrices are included when fitting a model in *mash*, *mash* will arbitrarily assign effect loadings across these matrices, as all assignments will lead to the same fit to the effect-size data. Our GxWeather covariance matrices were in many cases highly correlated, as they represented covariation in the same weather variables in overlapping time frames. We thus evaluated the performance of *mash* models when different covariance matrices were included by implementing a "greedy" mash algorithm.

Specifically, given $k$ covariance matrices for each phenotype, and $m$ corresponding effect sizes and standard errors across $n$ contexts of interest, we first fit a set of mash models with each of the $k$ matrices separately. We then identify the model with the maximum log likelihood estimates and select the corresponding matrix, leaving $k-1$ covariance matrices. We then fit a new set of mash models where the selected covariance matrix is paired with each of the remaining $k-1$ covariance matrices. The most likely pair of covariance matrices are then selected and the process is repeated for all possible $k-2$ matrix triplets. This process is repeated until one of two stop conditions are met: (i) all covariance matrices have been added to the mash model, resulting in a final model with all of the original $k$ covariance matrices are included or (ii) a likelihood ratio test with one degree of freedom between the likelihood of the current model and the model from the previous iteration has a *p*-value \> 0.05. The result is a set of log likelihood estimates for each model and the stepwise ‘path’ of covariance matrices that best explain the observed effect sizes and standard errors observed for the phenotype of interest.

We applied the greedy *mash* algorithm to the entire set of canonical, data-driven, and GxWeather covariance matrices described in Section S1. We used the random subset of SNP effect estimates, described in Section S3, in the greedy *mash* algorithm in order to test covariance matrices on an unbiased representation of all effects across the genome. The code to replicate this analysis can be found at git\@github.com:Alice-MacQueen/pvdiv-phenology-gxe.git/analysis/mash_greedy_algorithm/mash_h2_diag_covar/ in four directories: all_inputs/, gulf_inputs/, midwest_flowering/, and midwest_greenup/. For our six models of two phenological traits in three genetic subpopulations, between eight and twelve iterations of the greedy algorithm were necessary before a stop condition was met (@tbl-greedy). The relatively small number of necessary iterations, and correspondingly small number of covariance matrices selected, indicate that the hypothesis covariance matrices were highly correlated with one another and that inclusion of all the hypothesis covariance matrices introduced redundancy in the model. Most notably, the canonical "Identity" covariance matrix, representing independent effects, was the first matrix selected in all six applications of the greedy algorithm (@tbl-greedy). This indicated that a large number of independent effects were present in our analyses of both phenological traits. In addition, most or all of the available canonical covariance matrices, which specifed garden-specific effects, were selected before any GxWeather covariance matrices or data-driven covariance matrices (@tbl-greedy). Finally, at least one GxWeather covariance matrix was selected in each greedy algorithm application; in contrast, only one data-driven covariance matrix was selected (@tbl-greedy).

@@

To evaluate the prevalence and kinds of covariance patterns of SNP effects across our eight common gardens, we used multivariate adaptive shrinkage (*mash*) to jointly re-estimate SNP effect estimates for our strong effects subset using the set of covariance matrices selected by the greedy *mash* algorithm. Mash is a statistical method that allows estimation and comparison of many effects jointly across many different conditions; it improves on previous methods by allowing arbitrary correlations in effect sizes among conditions. Mash assigns mixture proportions for each SNP's effect onto each provided covariance matrix using maximum likelihood. Then, mash uses Bayes' theorem to shrink effects for each SNP towards the set of covariance matrices in accordance to their mixture proportions. The ability to specify multiple types of covariance matrix is an important advantage *mash* offers for studying patterns of GxE: user-specified covariance matrices allow hypothesis testing of specific environmental drivers for each SNP's effect, while the data-driven covariance matrices mash creates allow exploration of additional unexplained patterns of covariation.

The code used to run *mash* can be found at git\@github.com:Alice-MacQueen/pvdiv-phenology-gxe.git/R/Analysis_v1.2_mash_using_greedy_covar.qmd, starting at line 1325.

@@

| Phenology type | Subpopulation | Effects Selected per GWAS, Strong Subset | Effects Selected per GWAS, Random Subset | Total Markers, Strong Subset | Total Markers, Random Subset |
|:-----------|-----------:|-----------:|-----------:|------------|------------|
| Greenup Date   |          Gulf |                                    19000 |                                    38000 | 45931                        | 91862                        |
|                |       Midwest |                                    24000 |                                    48000 | 49030                        | 98060                        |
|                |          Both |                                    19000 |                                    38000 | 44765                        | 89530                        |
| Flowering Date |          Gulf |                                    19000 |                                    38000 | 50490                        | 100980                       |
|                |       Midwest |                                    33000 |                                    66000 | 63367                        | 126734                       |
|                |          Both |                                    19000 |                                    38000 | 35735                        | 71470                        |

@@

| Greedy Algorithm Iteration \# |     Gulf Greenup Date |           Midwest Greenup Date |    Both Greenup Date | Gulf Flowering Date     | Midwest Flowering Date         | Both Flowering Date     |
|:----------|----------:|----------:|-----------:|------------|------------|------------------|
| 1                             |              Identity |                       Identity |             Identity | Identity                | Identity                       | Identity                |
| 2                             |         Simple Het. 2 |               Single Effect MO |     Single Effect MI | Simple Het. 1           | Simple Het. 1                  | Single Effect TX3       |
| 3                             |      Single Effect MO |               Single Effect NE |     Single Effect SD | Single Effect MI        | Single Effect MI               | Single Effect TX1       |
| 4                             |      Single Effect NE |              Single Effect TX2 |     Single Effect NE | Single Effect TX3       | Single Effect TX2              | Single Effect OK        |
| 5                             | Daylength (14d prior) |              Single Effect TX1 |    Single Effect TX2 | Single Effect SD        | Single Effect MO               | Single Effect MI        |
| 6                             |      Single Effect MI |               Single Effect MI |     Single Effect OK | Single Effect OK        | Single Effect TX3              | Single Effect NE        |
| 7                             |      Single Effect SD |               Single Effect OK |    Single Effect TX3 | Single Effect NE        | Daylength change (s), 2d prior | Single Effect TX2       |
| 8                             |     Single Effect TX2 |              Single Effect TX3 |        Simple Het. 1 | Single Effect TX1       | Single Effect SD               | Single Effect MO        |
| 9                             |    Cumulative GDD, 6d |                  Simple Het. 1 |    Single Effect TX1 | Single Effect TX2       | \-                             | Single Effect SD        |
| 10                            |     Single Effect TX3 | Daylength change (s), 7d prior |     Single Effect MO | Single Effect MO        | \-                             | Simple Het. 1           |
| 11                            |   Data-driven $PCA_2$ |                             \- |         Temp Ave, 1d | Cumulative rainfall, 7d | \-                             | Cumulative rainfall, 1d |
| 12                            |                    \- |                             \- | Daylength change (s) | \-                      | \-                             | \-                      |

>>>>

## Section S1. Generation of covariance matrices to provide multiple ways SNP effects could covary across common gardens {.unnumbered}

A trait measured in two environments can be considered two traits that are genetically correlated [@falconer1952]. By extension, a phenological trait measured at eight common gardens can be modeled as eight genetically correlated traits represented by a covariance matrix, where diagonal entries represent the additive genetic variance present in the population at that garden, and non-diagonal entries represent the genetic correlation between the trait in a pair of environments.

We were interested in the types of genetic correlation that could be expressed by different alleles across the genome and reasoned that different loci could have different patterns of genetic correlation. For example, the effects of an allele of locus $A$ could covary with a weather-based cue if $A$ had a common allele, $a$, that was responsive to this cue. In contrast, the effects of a different locus $B$ might have an effect at only one garden, for example if a pathogen was only present at one garden, and $B$ has a common allele $b$ that is resistant to that pathogen.

@@

To create the GxWeather covariance matrices we transformed each genotype's green-up and flowering date at each garden to create date-related values that were summary functions for a weather variable across a specified date range. For example, the cumulative sum of the rainfall that occurred seven days prior to the flowering date was used to create one possible GxWeather cue variable for flowering date. The code for these calculations can be found at git\@github.com:Alice-MacQueen/pvdiv-phenology-gxe/R/Analysis_v1.2_mash_using_greedy_covar.qmd, starting at line 394.

@@

We then generated GxWeather covariance matrices derived from correlations in these GxWeather cue variables for the three subpopulations. These covariance matrices represent the correlations between genotypes for these weather-related variables across our eight common gardens. To create these matrices, we determined the correlations between these weather-related cue variables for identical genotypes grown at different common gardens and used these correlations to fill the off-diagonals of a eight-by-eight matrix. To fill the diagonal entries of these matrices, we used the narrow-sense heritability for each weather-related cue variable at each garden, calculated using the same methods as Section S3. Matrices need to be positive semi-definite to be used in *mash*, so when the resultant matrices were not positive semi-definite, the `Matrix::nearPD` function was used to compute the nearest positive definite matrix to the correlation matrix [@Matrix]. Finally, all covariance matrices were rescaled so that the maximum diagonal element of the rescaled matrix is 1, as in [@Urbut2019]. Code used to generate these matrices can be found at git\@github.com:Alice-MacQueen/pvdiv-phenology-gxe.git/R/Analysis_v1.2_mash_using_greedy_covar.qmd, starting at line 720.

@@

To obtain initial effect estimates for each single nucleotide polymorphism (SNP) for each combination of phenological trait, genetic subpopulation, and common garden, we conducted univarate genome-wide association (GWAS) on site-specific best linear unbiased predictors (BLUPs) using the switchgrassGWAS package and the methods in [@Lovell2021]. The full SNP diversity panel generated in [@Lovell2021] was subset within each subpopulation for all analyses to retain only SNPs with $\leq$ 20% missing data and minor allele frequencies \> 0.05. This subsetting resulted in 8.8 million SNPs retained for the Midwest subpopulation, 10.3 million SNPs retained for the Gulf subpopulation, and 12.3 million SNPs retained for Both subpopulations.

@@

where the vector $\mathbf{y}$ represents the flowering date or greenup date values for that garden, $Z$ the design matrix for random effects, $u$ the whole genome additive genetic effect, $e$ the residual, $G$ the genomic relationship matrix based on all SNPs retained for subpopulation-specific analyses, and $I$ the rank-$\mathbf{y}$ identity matrix [@clark2013]. ASReml can specify considerably more complex variance structures; however, we were interested in specifying covariance structures using *mash*, which allows the specification of multiple covariance structures; hence, we used a relatively simple model to calculate BLUPs for use in univariate GWAS. The code for these calculations can be found at git\@github.com:Alice-MacQueen/pvdiv-phenology-gxe.git/R/Analysis_v1.2_mash_using_greedy_covar.qmd, starting at line 75.

Univariate GWAS were run independently for each combination of trait, genetic subpopulation, and common garden. The code used to run GWAS can be found at git\@github.com:Alice-MacQueen/pvdiv-phenology-gxe.git/R/Analysis_v1.2_mash_using_greedy_covar.qmd, starting at line 240. We used [`pvdiv_standard_run()`](https://alice-macqueen.github.io/switchgrassGWAS/reference/pvdiv_standard_gwas.html), an opinionated function of the [switchgrassGWAS](https://github.com/Alice-MacQueen/switchgrassGWAS) R package, to conduct GWAS. To control for population structure and run GWAS, this function uses a three-part process. First, a fast truncated singular value decomposition (SVD) is generated from the SNP data, which has initial pruning of regions in long-range linkage disequilibrium (LD) and iterative pruning and removal of regions in long-range LD [@privé2017a]. By default, 15 vectors of singular values, $k$, are computed. Pruning and removal of long-range LD regions are recommended in research in human genetics to best control for population structure [@abdellaoui2013; @privé2017; @price2008]. Second, linear regression models using 0-15 of these vectors of singular values were used to run GWAS and to calculate the genomic control coefficient, $\lambda_{GC}$, for each $k$ [@Devlin1999]. Third, we chose the lowest value of $k$ that made $\lambda_{GC}$ closest to 1, and ran GWAS using a linear regression model with this $k$ for population structure correction.

@@

To make the subsequent analyses using *mash* computationally feasible, we followed the author's [recommended analysis outline](https://stephenslab.github.io/mashr/articles/eQTL_outline.html) for large datasets using *mash* and considered subsets of effects at different stages of the mash analysis. We extracted two subsets of tests from our complete GWAS effect estimate datasets: 1) effects from a subset of "strong" tests corresponding to stronger effects on our traits; 2) results from a random subset of all tests to correspond to an unbiased representation of all effects. We also clumped effects before selection of both subsets so that signal strength was standardized across different LD blocks in the genome, and not limited to closely linked associations.

@@

## Section S4. Greedy *mash* algorithm to select covariance matrices that significantly improved the model log-likelihood {.unnumbered}

Because we were particularly interested in determining whether genomic loci varied in their patterns of genetic covariance across gardens, and thus loaded on to different covariance matrices in *mash*, specifying the optimal set of covariance matrices in *mash* was an important model-building problem. If very similar covariance matrices are included when fitting a model in *mash*, *mash* will arbitrarily assign effect loadings across these matrices, as all assignments will lead to the same fit to the effect-size data. Our GxWeather covariance matrices were in many cases highly correlated, as they represented covariation in the same weather variables in overlapping time frames. We thus evaluated the performance of *mash* models when different covariance matrices were included by implementing a "greedy" mash algorithm.

Specifically, given $k$ covariance matrices for each phenotype, and $m$ corresponding effect sizes and standard errors across $n$ contexts of interest, we first fit a set of mash models with each of the $k$ matrices separately. We then identified the model with the maximum log likelihood estimates and selected the corresponding matrix, leaving $k-1$ covariance matrices. We then fit a new set of mash models where the selected covariance matrix is paired with each of the remaining $k-1$ covariance matrices. The most likely pair of covariance matrices was then selected and the process was repeated for all possible $k-2$ matrix triplets. This process was repeated until one of two stop conditions was met: (i) all covariance matrices have been added to the mash model, resulting in a final model with all of the original $k$ covariance matrices are included or (ii) a likelihood ratio test with one degree of freedom between the likelihood of the current model and the model from the previous iteration has a *p*-value \> 0.05. The result is a set of log likelihood estimates for each model and the stepwise ‘path’ of covariance matrices that best explain the observed effect sizes and standard errors observed for the phenotype of interest.

We applied the greedy *mash* algorithm to the entire set of canonical, data-driven, and GxWeather covariance matrices described in Section S1. We used the random subset of SNP effect estimates, described in Section S3, in the greedy *mash* algorithm in order to test covariance matrices on an unbiased representation of all effects across the genome. The code to replicate this analysis can be found at git\@github.com:Alice-MacQueen/pvdiv-phenology-gxe.git/analysis/mash_greedy_algorithm/mash_h2_diag_covar/ in four directories: all_inputs/, gulf_inputs/, midwest_flowering/, and midwest_greenup/. For our six models of two phenological traits in three genetic subpopulations, between eight and twelve iterations of the greedy algorithm were necessary before a stop condition was met (@tbl-greedy). The relatively small number of necessary iterations, and correspondingly small number of covariance matrices selected, indicate that the hypothesis covariance matrices were highly correlated with one another and that inclusion of all the hypothesis covariance matrices would introduce redundancy into the model. Most notably, the canonical "Identity" covariance matrix, representing independent effects, was the first matrix selected in all six applications of the greedy algorithm (@tbl-greedy). This indicated that a large number of independent effects were present in our analyses of both phenological traits. In addition, most or all of the available canonical covariance matrices, which specifed garden-specific effects, were selected before any GxWeather covariance matrices or data-driven covariance matrices (@tbl-greedy). Finally, at least one GxWeather covariance matrix was selected in each greedy algorithm application; in contrast, one data-driven covariance matrix was selected in just one of the six models (@tbl-greedy).

@@

To evaluate the prevalence and kinds of covariance patterns of SNP effects across our eight common gardens, we used multivariate adaptive shrinkage (*mash*) to jointly re-estimate SNP effect estimates for our strong effects subset using the set of covariance matrices selected by the greedy *mash* algorithm. *Mash* is a statistical method that allows estimation and comparison of many effects jointly across many different conditions; it improves on previous methods by allowing arbitrary correlations in effect sizes among conditions. *Mash* assigns mixture proportions for each SNP's effect onto each provided covariance matrix using maximum likelihood. Then, *mash* uses Bayes' theorem to shrink effects for each SNP towards the set of covariance matrices in accordance to their mixture proportions. The ability to specify multiple types of covariance matrix is an important advantage *mash* offers for studying patterns of GxE: user-specified covariance matrices allow hypothesis testing of specific drivers for each SNP's effect, such as weather-related cues, while the data-driven covariance matrices *mash* creates allow exploration of additional unexplained patterns of covariation.

The code used to run *mash* can be found at git\@github.com:Alice-MacQueen/pvdiv-phenology-gxe.git/R/Analysis_v1.2_mash_using_greedy_covar.qmd, starting at line 1326.

@@

| Phenology type | Subpopulation | Effects Selected per GWAS, Strong Subset | Effects Selected per GWAS, Random Subset | Total Markers, Strong Subset | Total Markers, Random Subset |
|:---------------|--------------:|-----------------------------------------:|-----------------------------------------:|------------------------------|------------------------------|
| Greenup Date   |          Gulf |                                    19000 |                                    38000 | 45931                        | 91862                        |
|                |       Midwest |                                    24000 |                                    48000 | 49030                        | 98060                        |
|                |          Both |                                    19000 |                                    38000 | 44765                        | 89530                        |
| Flowering Date |          Gulf |                                    19000 |                                    38000 | 50490                        | 100980                       |
|                |       Midwest |                                    33000 |                                    66000 | 63367                        | 126734                       |
|                |          Both |                                    19000 |                                    38000 | 35735                        | 71470                        |

@@

| Greedy Algorithm Iteration \# |     Gulf Greenup Date |           Midwest Greenup Date |    Both Greenup Date | Gulf Flowering Date     | Midwest Flowering Date         | Both Flowering Date     |
|:------------------------------|----------------------:|-------------------------------:|---------------------:|-------------------------|--------------------------------|-------------------------|
| 1                             |              Identity |                       Identity |             Identity | Identity                | Identity                       | Identity                |
| 2                             |         Simple Het. 2 |               Single Effect MO |     Single Effect MI | Simple Het. 1           | Simple Het. 1                  | Single Effect TX3       |
| 3                             |      Single Effect MO |               Single Effect NE |     Single Effect SD | Single Effect MI        | Single Effect MI               | Single Effect TX1       |
| 4                             |      Single Effect NE |              Single Effect TX2 |     Single Effect NE | Single Effect TX3       | Single Effect TX2              | Single Effect OK        |
| 5                             | Daylength (14d prior) |              Single Effect TX1 |    Single Effect TX2 | Single Effect SD        | Single Effect MO               | Single Effect MI        |
| 6                             |      Single Effect MI |               Single Effect MI |     Single Effect OK | Single Effect OK        | Single Effect TX3              | Single Effect NE        |
| 7                             |      Single Effect SD |               Single Effect OK |    Single Effect TX3 | Single Effect NE        | Daylength change (s), 2d prior | Single Effect TX2       |
| 8                             |     Single Effect TX2 |              Single Effect TX3 |        Simple Het. 1 | Single Effect TX1       | Single Effect SD               | Single Effect MO        |
| 9                             |    Cumulative GDD, 6d |                  Simple Het. 1 |    Single Effect TX1 | Single Effect TX2       | \-                             | Single Effect SD        |
| 10                            |     Single Effect TX3 | Daylength change (s), 7d prior |     Single Effect MO | Single Effect MO        | \-                             | Simple Het. 1           |
| 11                            |   Data-driven $PCA_2$ |                             \- |         Temp Ave, 1d | Cumulative rainfall, 7d | \-                             | Cumulative rainfall, 1d |
| 12                            |                    \- |                             \- | Daylength change (s) | \-                      | \-                             | \-                      |
